---
title: "AMAK (BSAI Atka mackerel)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{AMAK (BSAI Atka mackerel)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Getting started with `{afscOSA}`

To get started you'll need to install `{afscOSA}`, which also relies on `{compResidual}`. To install these libraries from Github, using the following commands:

```{r setup}

# downloading compResidual:
# https://github.com/fishfollower/compResidual#composition-residuals for
# installation instructions

# TMB:::install.contrib("https://github.com/vtrijoulet/OSA_multivariate_dists/archive/main.zip")
# remotes::install_github("fishfollower/compResidual/compResidual", force=TRUE)

# remotes::install_github("noaa-afsc/afscOSA", force=TRUE)

library(afscOSA)
```
In this vignette we show how to `{afscOSA}` with an AMAK ADMB model using the BSAI Atka mackerel assessment model as an example. 

Once the data is loaded, the general workflow is as follows:

1.  Structure the observed and predicted age/length compositions as matrices with `nrows` = number of years and `ncols` = number of age or length bins. 

2.  Calculate OSA residuals for each fleet using `run_osa()`. See details for inputs and outputs by running `??run_osa()`.

3.  Plot OSA residuals and aggregate fits for one or more fits using `plot_osa()`. Input to `plot_osa()` is a list of output(s) from `run_osa()`. See more details by running `??plot_osa`.

```{r,  fig.width = 6, fig.height = 8}
# load Atka mackerel data
amrep <- afscOSA::bsaiamrep
amdat <- afscOSA::bsaiamdat

# fishery
ages <- 1:11
yrs <- amrep$pobs_fsh_1[,1]
obs <- amrep$pobs_fsh_1[,2:12]
exp <- amrep$phat_fsh_1[,2:12]
N <- amdat$sample_ages_fsh
out1 <- run_osa(fleet = 'Fishery', index_label = 'Age',
                obs = obs, exp = exp, N = N, index = ages, years = yrs)

# survey
yrs <-  amrep$pobs_ind_1[,1]
obs <- amrep$pobs_ind_1[,2:12]
exp <- amrep$phat_ind_1[,2:12]
N <- amdat$sample_ages_ind
out2 <- run_osa(fleet = 'AI Trawl Survey', index_label = 'Age',
                obs = obs, exp = exp, N = N, index = ages, years = yrs)

input <- list(out1, out2)
osaplots <- plot_osa(input)
```
