---
title: "SS3 (AI Pacific cod)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SS3 (AI Pacific cod)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Getting started with `{afscOSA}`

To get started you'll need to install `{afscOSA}`, which also relies on `{compResidual}`. To install these libraries from Github, using the following commands:

```{r setup}

# downloading compResidual:
# https://github.com/fishfollower/compResidual#composition-residuals for
# installation instructions

# TMB:::install.contrib("https://github.com/vtrijoulet/OSA_multivariate_dists/archive/main.zip")
# remotes::install_github("fishfollower/compResidual/compResidual", force=TRUE)

# remotes::install_github("noaa-afsc/afscOSA", force=TRUE)

library(afscOSA)

```
In this vignette we show how to `{afscOSA}` with a Stock Synthesis (SS3) model using the Aleutian Islands Pacific cod assessment model as an example. For that reason, you'll also also need the `{r4ss}` library.

```{r r4ss}


# remotes::install_github("r4ss/r4ss", force=TRUE)

library(r4ss)

```


Once the data is loaded, the general workflow is as follows:

1.  Structure the observed and predicted age/length compositions as matrices with `nrows` = number of years and `ncols` = number of age or length bins. 

2.  Calculate OSA residuals for each fleet using `run_osa()`. See details for inputs and outputs by running `??run_osa()`.

3.  Plot OSA residuals and aggregate fits for one or more fits using `plot_osa()`. Input to `plot_osa()` is a list of output(s) from `run_osa()`. See more details by running `??plot_osa`.

```{r,  fig.width = 10, fig.height = 12}

sx = 1 # USER INPUT define sex
fleet = c(1,2) # USER INPUT define fleets

# use dirvec argument to point to the directory with your SS3 output files
mod <- r4ss::SSgetoutput()

# comps for the fleets defined in "fleet" and "sx"
comps <- as.data.frame(mod[[1]]$lendbase[,c(1,6,13,16:18)])
comps <- comps[comps$Fleet %in% fleet & comps$Sex %in% sx, ]
comps <- reshape2::melt(comps,id.vars = c('Yr','Fleet','Sex','Bin'))

# input sample sizes for the fleets defined in "fleet" and "sx"
Ndf <- as.data.frame(mod[[1]]$lendbase[,c(1,6,13,16,22)])
Ndf <- Ndf[Ndf$Bin == min(Ndf$Bin),]

# length bins
lens <- sort(unique(comps$Bin))

# fishery (fleet 1) ----

flt <- 1 # USER INPUT

# this is a 1 sex model but if it had sex structure you would define each sex as
# a separate fleet (e.g., Fishery F and Fishery M)
tmp <- comps[comps$Fleet==flt,]

# input sample sizes (vector)
N <- Ndf$effN[Ndf$Fleet==flt]

# observed values -> put in matrix format (nrow = nyr, ncol = age/length)
obs <- tmp[tmp$variable=='Obs',]
obs <- reshape2::dcast(obs, Yr~Bin, value.var = "value")
yrs <- obs$Yr # years sampled
obs <- as.matrix(obs[,-1])

# expected values -> put in matrix format (nrow = nyr, ncol = age/length
exp <- tmp[tmp$variable=='Exp',]
exp <- reshape2::dcast(exp, Yr~Bin, value.var = "value")
exp <- as.matrix(exp[,-1])

# should all be true!
length(N) == length(yrs); length(N) == nrow(obs); nrow(obs) == nrow(exp)
ncol(obs);ncol(exp);length(lens)

out1 <- afscOSA::run_osa(fleet = 'Fishery', index_label = 'Length',
                         obs = obs, exp = exp, N = N, index = lens, years = yrs)


# AI bottom trawl survey (fleet 2) ----

flt <- 2 # USER INPUT

# this is a 1 sex model but if it had sex structure you would define each sex as
# a separate fleet (e.g., Survey F and Survey M)
tmp <- comps[comps$Fleet==flt,]

# input sample sizes (vector)
N <- Ndf$effN[Ndf$Fleet==flt]

# observed values -> put in matrix format (nrow = nyr, ncol = age/length)
obs <- tmp[tmp$variable=='Obs',]
obs <- reshape2::dcast(obs, Yr~Bin, value.var = "value")
yrs <- obs$Yr # years sampled
obs <- as.matrix(obs[,-1])

# expected values -> put in matrix format (nrow = nyr, ncol = age/length
exp <- tmp[tmp$variable=='Exp',]
exp <- reshape2::dcast(exp, Yr~Bin, value.var = "value")
exp <- as.matrix(exp[,-1])

# should all be true!
length(N) == length(yrs); length(N) == nrow(obs); nrow(obs) == nrow(exp)
ncol(obs);ncol(exp);length(lens)

out2 <- afscOSA::run_osa(fleet = 'AI Trawl Survey', index_label = 'Length',
                         obs = obs, exp = exp, N = N, index = lens, years = yrs)

# plot results ----
input <- list(out1, out2)
osaplots <- plot_osa(input, use_agg_proportions = TRUE)

```

Note on Residual Truncation:

The warning above indicates that extreme Pearson residuals (e.g., 8.05, 12.59)
were capped at 6 for visualization. Truncating extreme outliers prevents them
from compressing the overall scale, allowing residual patterns across age bins
and years to remain clearly legible and directly comparable across different
assessment models.

There are a lot of length bins in this Pacific cod example, making examination of individual residuals challenging. To help with this, the user can extract the different plotting components:

```{r,  fig.width = 10, fig.height = 12}
osaplots <- plot_osa(input, use_agg_proportions=TRUE, plot=FALSE)
osaplots$bubble
# osaplots$bubble_pearson
```

```{r,  fig.width = 10, fig.height = 6}
osaplots <- plot_osa(input, use_agg_proportions=TRUE, plot=FALSE)
osaplots$qq
osaplots$aggcomp
```
